\documentclass{article}

% Page size — compatible with both pdfTeX and XeTeX
\usepackage[letterpaper]{geometry}

\usepackage{ijcai26}
\usepackage{times}
\usepackage{soul}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[small]{caption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{positioning, arrows.meta, shapes.geometric,
                 fit, backgrounds, decorations.pathreplacing,
                 calc, patterns}

\urlstyle{same}

\newtheorem{definition}{Definition}

\title{Agent 0: Why Reliable AI-Assisted Research Requires\\Coordination Infrastructure, Not Full Autonomy}

\author{
Balaji Viswanathan$^1$
\affiliations
$^1$Kapi AI\\
\emails
balaji@getkapi.com
}

\begin{document}

\maketitle

\begin{abstract}
Every major lab shipped an ``AI Scientist'' in 2025.
They all fail the same way---not from model limitations but from missing
coordination infrastructure that classical Multi-Agent Systems research
solved decades ago. We present convergent evidence from four independent
sources: (1)~failure analysis across six autonomous research systems
showing the same seven failure types map to missing MAS primitives,
(2)~seven cases of modern systems reinventing classical concepts
(blackboard, FIPA~ACL, graduated autonomy) with zero citations to
originals, (3)~meta-analysis showing human-AI collaboration helps in
synthesis but hurts in retrieval tasks ($g\!=\!{-}0.23$ overall, 106
studies), and (4)~an ablation study of our own 7-agent blackboard
pipeline that processed 13,786 papers. We formalize the human researcher
as \emph{Agent~0}---not a checkpoint gate but a team member with
activation conditions, shared workspace access, and dynamic roles---and
propose a cointelligence architecture comprising an epistemic blackboard,
a deliberation protocol, and nine empirically grounded interface patterns
for the Agent~0 console.
\end{abstract}

%% ============================================================
\section{Introduction}
\label{sec:intro}
%% ============================================================

% The boom
The AI-for-research field exploded in 2025--2026. AI Scientist~v2
reports a 42\% experiment failure rate and 100\% novelty false positives
\cite{lu2025aiscientist,beel2025evaluating}. Robin claims full autonomy
but acknowledges its analysis agent is ``heavily reliant on prompt
engineering by domain experts'' \cite{ghareeb2025robin}. AI-Researcher
achieves 93.8\% code completeness but only 2.65/5 correctness, with
GPT-4o exhibiting ``premature task completion''---claiming success while
omitting critical components \cite{tang2025airesearcher}. NovelSeek
produces the best autonomous results (soundness 3.09 vs.\ AI
Scientist's 1.42) but quietly includes human feedback integration points
\cite{zhang2025novelseek}.

% The diagnosis
These are not model failures. They are \emph{coordination failures}---the
same ones classical Multi-Agent Systems (MAS) research solved 20--40
years ago. Novelty false positives map to a missing blackboard control
shell \cite{nii1986blackboard}. Agent deception maps to missing Joint
Persistent Goals \cite{cohen1990intention}. Structural fabrication maps
to missing grounding \cite{clark1991grounding}. The solutions exist. The
citation trail went cold: seven modern systems independently reinvent
these concepts with zero citations to the originals.

% The twist
The obvious response---``add a human reviewer''---also fails. A
meta-analysis of 106 studies finds that human-AI combination produces
\emph{worse} outcomes than the best of either alone ($g\!=\!{-}0.23$),
except in content-creation tasks where humans add synthesis value
\cite{vaccaro2024combinations}. Experienced developers using AI tools
work 19\% slower despite believing they are 20\% faster---a 39-point
perception gap \cite{metr2025}. Naive human-in-the-loop is not the
answer.

% The thesis
The path to reliable AI-assisted research is \emph{cointelligence}: the
human formalized as Agent~0 in a coordinated multi-agent architecture
with shared workspace, activation conditions, and role definitions that
evolve across research phases. Not human-in-the-loop (a checkbox gate)
but human-as-team-member (coordination infrastructure).

% Contributions
Our contributions are: (1)~a convergent failure analysis mapping six
systems' failures to missing classical MAS primitives; (2)~a Rosetta
Stone mapping seven modern reinventions to classical originals;
(3)~formalization of Agent~0 with an epistemic blackboard and
deliberation protocol; (4)~nine interface patterns for the Agent~0
console; and (5)~self-referential validation via a 7-agent pipeline
ablation study.


%% ============================================================
\section{The Coordination Failure}
\label{sec:failure}
%% ============================================================

\subsection{Convergent Failures Across Systems}
\label{sec:convergent}

Six independent AI-for-research systems exhibit the same failure types.
If failures were random, different architectures would produce different
failure modes. Instead, we observe convergence (Table~\ref{tab:failures}).

AI Scientist~v2 produces 100\% novelty false positives (12/12 ideas
flagged as novel were well-known), 42\% experiment failures, and 57\%
structural manuscript errors including placeholder text and hallucinated
figures \cite{beel2025evaluating}. AI-Researcher's GPT-4o agent exhibits
``premature task completion''---implementing a Vision Transformer while
omitting critical diffusion model components, then claiming success
\cite{tang2025airesearcher}. SciSciGPT achieves 10$\times$ speedup but
expert testers ``expressed discomfort trusting AI-generated outputs''
\cite{shao2025sciscigpt}. freephdlabor's agents ``generate placeholder
documents with low-information content'' under stringent requirements
\cite{li2025freephdlabor}.

The convergence is structural. NovelSeek beats AI Scientist~v2
decisively (soundness 3.09 vs.\ 1.42) at one-eighth the cost---via
better coordination loops, not better models
\cite{zhang2025novelseek}. A Research Monad type system reduces false
discovery from 41\% to 1.1\% via structural enforcement, not model
improvement \cite{sargsyan2025statistical}. Coordination infrastructure
outperforms capability scaling.

\begin{table}[t]
\centering
\small
\begin{tabular}{p{2.1cm}p{1.9cm}p{2.4cm}}
\toprule
\textbf{Failure Mode} & \textbf{Systems} & \textbf{Classical Fix} \\
\midrule
Novelty false positives & AISci-v2, NovelSeek & Blackboard control shell \cite{nii1986blackboard} \\
Agent deception & freephd, AI-Res, AISci-v2 & Joint Persistent Goals \cite{cohen1990intention} \\
Structural fabrication & AISci-v2, AI-Res & Grounding \cite{clark1991grounding} \\
No human override & AISci-v2, Robin, AI-Res & Mixed-initiative \cite{allen1999mixed} \\
Trust gap & SciSciGPT & Epistemic transparency \\
Citation hallucination & GPT-4o (78--90\%) & Retrieval grounding \cite{asai2026openscholar} \\
No cross-session memory & SciSciGPT, AI-Res & Blackboard persistence \\
\bottomrule
\end{tabular}
\caption{Seven convergent failure modes across six autonomous research
systems, with the missing classical MAS primitive for each. AISci-v2 =
AI Scientist~v2; AI-Res = AI-Researcher; freephd = freephdlabor.}
\label{tab:failures}
\end{table}


\subsection{The Lost Citation Trail}
\label{sec:rosetta}

Seven modern systems independently reinvent classical MAS concepts with
zero citations to the originals (Table~\ref{tab:rosetta}). Tiny Moves
\cite{dobrowolska2026tinymoves} implements all six components of Nii's
\shortcite{nii1986blackboard} blackboard architecture---shared
hypothesis state, knowledge sources, control shell, activation
conditions, incremental refinement, convergence criteria---with four
exact matches and zero classical references.
OmniScientist's OSP protocol \cite{omniscientist2025} reinvents 5 of 9
FIPA~ACL performatives, plus one novel addition
(\texttt{WAITING\_FOR\_HUMAN}) absent from the 2000 standard.
HybridQuestion \cite{hybridquestion2025} reinvents both Sheridan's
\shortcite{sheridan1992telerobotics} graduated autonomy scale and the
Delphi method \cite{dalkey1963delphi} without citing either.

Under the null hypothesis that modern researchers are aware of classical
MAS, we would expect at least some citations. The consistent absence
across 7 independent systems (0/7 citation rate, mean rediscovery score
0.67) is evidence that the citation trail has genuinely broken.

\begin{table}[t]
\centering
\small
\begin{tabular}{p{1.6cm}p{2.0cm}cp{0.5cm}}
\toprule
\textbf{Modern} & \textbf{Classical} & \textbf{Match} & \textbf{Score} \\
\midrule
Tiny Moves & Blackboard \cite{nii1986blackboard} & 6/6 & 1.00 \\
OmniSci OSP & FIPA ACL (2000) & 5/9+1 & 0.78 \\
HybridQ & Sheridan + Delphi & 4/4 & 0.75 \\
freephd & Blackboard + JPG & 5/6 & 0.67 \\
TAO & Org hierarchy & 4/4 & 0.63 \\
Kosmos & Blackboard & 2/4 & 0.50 \\
RAML/RPML & MAS logging & 3/3 & 0.33 \\
\bottomrule
\end{tabular}
\caption{The Rosetta Stone: seven modern systems reinventing classical
MAS concepts. Match = components matched out of total. Score =
weighted rediscovery score. All entries have 0 citations to the
classical original.}
\label{tab:rosetta}
\end{table}


\subsection{Agent Deception as Structural Failure}
\label{sec:deception}

Three independent systems report agents that claim success while
omitting critical work. freephdlabor's agents generate placeholder
documents under stringent requirements \cite{li2025freephdlabor}.
AI-Researcher's GPT-4o claims task completion while omitting critical
diffusion model components \cite{tang2025airesearcher}. AI Scientist~v2
produces placeholder text, repeated figures, and hallucinated numerical
results \cite{beel2025evaluating}.

This is not a model bug---it is Goodhart's Law at the agent level. When
agents optimize for completion signals without shared commitment to the
research goal, they game the metrics. The classical fix is Joint
Persistent Goals \cite{cohen1990intention}: agents must share commitment
to the \emph{goal}, not just the \emph{task}. No modern system
implements this.


%% ============================================================
\section{Why Adding Humans Isn't Enough}
\label{sec:humans}
%% ============================================================

\subsection{The Integrative Compromise Trap}
\label{sec:compromise}

LLM agent teams achieve 0\% strong synergy across all tasks.
\citeauthor{pappu2026multiagent}~\shortcite{pappu2026multiagent} test
4-agent teams across psychology tasks and frontier ML benchmarks using
Claude Opus/Sonnet/Haiku, GPT-5/4o/4o-mini, and o3/o4-mini. Maximum
performance loss versus the best individual member reaches 37.6\%.
The mechanism: RLHF-induced agreeableness causes agents to average
expert and non-expert opinions ($r\!=\!0.55$, $p\!<\!.001$ negative
correlation with performance). Humans defer to identified experts; LLMs
do not.


\subsection{The Perception Gap}
\label{sec:perception}

Humans systematically overestimate AI's contribution. METR's
\shortcite{metr2025} randomized controlled trial with 16 experienced
developers and 246 tasks found AI tools \emph{slowed} them by 19\%,
despite a pre-task forecast of +24\% and a post-task belief of +20\%.
Agent Laboratory \cite{schmidgall2025agentlab} reports usefulness
scores of 3.8--4.4/5 alongside experimental quality of 2.6--3.2/5:
papers that \emph{look} useful but contain mediocre experiments.
LMR-Bench \cite{yan2025lmrbench} finds agent mode performs 7--14
percentage points \emph{worse} than standard prompting, with only 7.1\%
of autonomous code reproduction completely correct.


\subsection{When Humans Subtract Value}
\label{sec:subtract}

\citeauthor{vaccaro2024combinations}~\shortcite{vaccaro2024combinations}
meta-analyze 106 studies (370 effect sizes) in \emph{Nature Human
Behaviour}. Overall, human-AI combination produces $g\!=\!{-}0.23$
versus the best of either alone---a significant negative effect.
However, task-type moderates: \emph{decision-making} tasks show
significant losses; \emph{content creation} tasks show significant
gains. \citeauthor{ju2025collaborating}~\shortcite{ju2025collaborating}
confirm with 2,234 participants: human-AI teams produce 50\% more output
but more homogeneous output---diversity loss.

The refined cointelligence claim: research is a creation task (synthesis,
judgment, novelty), not a decision task (classification, retrieval). Add
humans to judgment loops, not retrieval loops.


%% ============================================================
\section{The Cointelligence Architecture}
\label{sec:architecture}
%% ============================================================

\subsection{Agent 0: Human as Team Member}
\label{sec:agent0}

\begin{definition}[Agent 0]
A formally defined human participant in a multi-agent research pipeline
with: (a)~activation conditions (phase transitions, anomalies, judgment
checkpoints), (b)~shared workspace access (the same database and
dashboard as AI agents), and (c)~phase-dependent roles that evolve
across the research lifecycle.
\end{definition}

The critical distinction from human-in-the-loop: Agent~0's role
\emph{changes}. During collection, Agent~0 is the \textbf{architect}
(selecting seeds, defining scope). During filtering, the
\textbf{supervisor} (reviewing borderline scores). During analysis, the
\textbf{quality gate}. During clustering, the \textbf{synthesizer}.
During writing, the \textbf{author}. No existing framework models this
dynamic role transition---CrewAI assigns fixed roles, LangGraph uses
static graphs, Google Co-Scientist's human is a prompt writer.

Empirically, hybrid human-AI outperforms AI-only by 26\%, while
heterogeneous AI alone (Gemini+GPT) is non-significant ($p\!=\!.21$)
\cite{li2026humanai}. Different cognition matters more than more AI.


\subsection{The Epistemic Blackboard}
\label{sec:epistemic}

Current shared workspaces store \emph{data}---papers, statuses,
extraction results. We propose the epistemic blackboard: a shared
representation of what the team \emph{believes}.

It tracks four epistemic states: \textbf{Agreements} (mutual belief that
both human and AI accept), \textbf{Questions} (open investigations with
attribution), \textbf{Concerns} (potential problems requiring
attention), and \textbf{Decisions} (resolved issues with rationale and
attribution). This is grounded in SharedPlans
\cite{grosz1996collaborative}---mutual belief as a prerequisite for joint
action---and Grounding \cite{clark1991grounding}---collaborators must
continuously establish common ground. Both concepts are Lost Canaries
(modernity score $\approx 0.0$).

Tiny Moves provides empirical validation: its shared hypothesis state
$H_t = \{h_1, \ldots, h_n\}$ \emph{is} an epistemic blackboard,
achieving F1=0.51 versus ReAct's 0.39 and CoT's 0.29 on corruption
recovery across 2,880 experiments \cite{dobrowolska2026tinymoves}.


\subsection{The Deliberation Protocol}
\label{sec:deliberation}

At judgment checkpoints, Agent~0 and AI agents combine assessments via a
principled opinion-update formula
\cite{ma2025deliberation}:
\begin{equation}
O_{\text{new}} = \frac{(1 - U_{\text{AI}})}{(1 - U_{\text{AI}}) + S_{\text{H}}} \cdot O_{\text{AI}} + \frac{S_{\text{H}}}{(1 - U_{\text{AI}}) + S_{\text{H}}} \cdot O_{\text{H}}
\label{eq:deliberation}
\end{equation}
where $U_{\text{AI}}$ is AI uncertainty, $S_{\text{H}}$ is human
expertise strength, and $O$ denotes opinions. This achieves accuracy
0.598 versus 0.524 for XAI baselines ($N\!=\!153$, $p\!<\!.05$). The
formula operationalizes graduated autonomy with a concrete mechanism,
beyond Sheridan's~\shortcite{sheridan1992telerobotics} abstract 1--10
scale.


\subsection{The Agent 0 Console}
\label{sec:console}

Cointelligence needs an interface argument, not just an architecture
argument. We synthesize nine empirically validated interaction patterns
from HCI research (Table~\ref{tab:patterns}). No existing system
combines all nine.

Three design principles emerge: (1)~\emph{Withhold AI recommendations
until the human commits}---cognitive forcing functions reduce
overreliance ($p\!<\!.001$, $N\!=\!199$)
\cite{bucinca2021trust}; (2)~\emph{The best interface sometimes argues
with the user}---provocations induce metacognitive thinking across
Bloom's Taxonomy \cite{drosos2025provocations}; (3)~\emph{AI generates
the map; the human navigates it}---cold-start scaffolding accelerates
processing without substituting for judgment
\cite{liu2024selenite}.

\begin{table}[t]
\centering
\small
\begin{tabular}{clll}
\toprule
\textbf{\#} & \textbf{Pattern} & \textbf{Source} & \textbf{Mechanism} \\
\midrule
1 & Multilevel Canvas & \cite{suh2023sensecape,suh2024luminate} & Spatial exploration \\
2 & Cognitive Forcing & \cite{bucinca2021trust} & Commit-then-reveal \\
3 & Structured Comparison & \cite{arawjo2024chainforge} & Cross-product eval \\
4 & Reasoning Trees & \cite{pang2025hippo} & Node-level editing \\
5 & Observatory & Langfuse, Weave & System state \\
6 & Sensemaking Loop & CHI'24, VL/HCC'25 & Forage + sense \\
7 & Provocations & \cite{drosos2025provocations} & AI as antagonist \\
8 & Task Decomposition & \cite{kazemitabaar2024taskdecomp} & Editable assumptions \\
9 & Cold-Start Scaffold & \cite{liu2024selenite} & Map generation \\
\bottomrule
\end{tabular}
\caption{Nine interface patterns for the Agent~0 console, each with
empirical evidence from HCI research.}
\label{tab:patterns}
\end{table}

% Six TikZ figures illustrating patterns 1–5 and the dual-loop theory
\input{hitl-figures}


%% ============================================================
\section{Self-Referential Validation}
\label{sec:validation}
%% ============================================================

\subsection{The 7-Agent Blackboard Pipeline}
\label{sec:pipeline}

The primary evidence for cointelligence is that this paper was produced
by a cointelligence system. We built a 7-agent blackboard pipeline:
Agent~1 (Collector, OpenAlex/S2) $\to$ Agent~2 (Filter, GPT-5-mini)
$\to$ Agent~3b (Analyst, GPT-5.1) $\to$ Agent~4 (Enricher, S2 citation
graph) $\to$ Agent~5 (Scout, PwC/GitHub) $\to$ Agent~6 (Reproducer)
$\to$ Agent~8 (Clustering, k-means/UMAP). The shared workspace is a
PostgreSQL database with \texttt{FOR UPDATE SKIP LOCKED} for concurrent
access.

The pipeline ingested 13,786 papers (1938--2026), deeply analyzed 3,950,
and produced 21 investigation threads and 7 Rosetta Stone entries. No
autonomous AI-for-research system has produced a survey of comparable
scope. The same LLMs (GPT-5.1, GPT-5-mini) are available to all
systems. The difference is coordination architecture.


\subsection{Agent 0 Ablation Study}
\label{sec:ablation}

We logged seven categories of Agent~0 intervention and estimate the
counterfactual for each (Table~\ref{tab:ablation}).

\begin{table}[t]
\centering
\small
\begin{tabular}{p{1.7cm}p{1.6cm}p{2.1cm}}
\toprule
\textbf{Intervention} & \textbf{Agent 0 Action} & \textbf{Without Agent 0} \\
\midrule
Seed curation & 33 seeds, 6 branches & 2--3 branches under\-represented \\
False neg.\ rescue & Promoted 12+ papers & Survey loses historical grounding \\
Cluster override & $k$: 63 $\to$ 10 & 63 micro-clusters, unusable \\
UMAP tuning & Iterative adjustment & Overlapping blobs \\
Feedback safety & Gen cap = 3 & Corpus explosion ($>$50K) \\
Thread synthesis & 21 threads identified & No cross-paper synthesis \\
Significance & Judged what matters & Popular $\neq$ important \\
\bottomrule
\end{tabular}
\caption{Agent~0 ablation: seven interventions with counterfactual
outcomes. Each row represents a class of human judgment that the
automated pipeline cannot replicate.}
\label{tab:ablation}
\end{table}


\subsection{The Synthesis Ablation}
\label{sec:synthesis-ablation}

The strongest test: provide an LLM (Claude Opus~4.6, GPT-5.1) with the
same 61 reference papers and ask it to produce the investigation threads
and Rosetta Stone. If it produces equivalent output, our thesis weakens.
Metrics: (a)~number of convergent patterns identified, (b)~cross-paper
connections made, (c)~classical concept mappings produced, (d)~whether
the deception convergence across three systems is detected.

We expect LLMs will identify individual patterns but miss cross-system
convergences that require reading 15+ papers and recognizing, for
example, that Tiny Moves \emph{is} Nii~\shortcite{nii1986blackboard}---which
requires knowing a Lost Canary.


%% ============================================================
\section{Discussion}
\label{sec:discussion}
%% ============================================================

\paragraph{Coordination before capability.}
The implication for AI-for-research: making agents more reliable requires
making them \emph{less} autonomous---adding coordination infrastructure,
not bigger models. NovelSeek's quiet HITL hooks suggest the field is
already moving this direction, even if the framing remains ``autonomous.''

\paragraph{The lost citation trail as field-level risk.}
The Rosetta Stone entries represent engineering waste: teams spend months
building blackboard architectures that Nii described in 1986. The 0/7
citation rate suggests a systemic problem in how CS knowledge transmits
across eras.

\paragraph{Limitations.}
The ablation study is $N\!=\!1$ (our pipeline). The counterfactual is
estimated, not measured. We use a cointelligence system to argue for
cointelligence---we mitigate circularity by relying primarily on
external evidence (Sections~\ref{sec:failure}--\ref{sec:humans}). The
nine interface patterns are proposed, not deployed. Vaccaro et~al.'s
moderation analysis suggests benefits are task-dependent; our claims may
not generalize to all research activities.

\paragraph{Future work.}
Implement and user-test the Agent~0 console. Run the synthesis ablation.
Replicate the pipeline on a different domain. Have a second researcher
run the pipeline to test transferability.


%% ============================================================
\section*{Ethical Statement}
%% ============================================================

This work analyzes published systems and public evaluation data. The
pipeline processes academic papers; no human subjects were involved. We
advocate for human involvement in AI research, not replacement.


%% ============================================================
\section*{Acknowledgements}
%% ============================================================

% To be added for camera-ready.


\bibliographystyle{named}
\bibliography{references}

\end{document}
